---
title: "PA_healthinsurance"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
date: "2024-01-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r install, eval = FALSE}
# Run once if these packages aren't installed
install.packages(c("tidyverse", "corrplot", "hexbin", "e1071", "shiny"))
```

```{r}
library(tidyverse)
# data/health_insurance.csv is the same data as the original health_insurance.RData
data <- read.csv("data/health_insurance.csv") %>%
  mutate_if(is.character, as.factor)

data %>% 
  summary()

```
First, gather some basic data about the population: 

The total number of people in this dataset is 1338 with 662 females and 676 males. Our youngest individual is 18 years of age whereas our oldest individual is 64.
The max amount of children someone has is 5 whereas the lowest number of children one could have is 0.
There are 1064 people who are non-smokers and 274 who are smokers. 
The lowest BMI is 15.96 and the highest BMI is 53.13.
There are 324 from the Northeast region, 325 from the Northwest, 364 from the Southeast and 325 from the Southwest.
The average charge per individual is 13,270.

If we wanted, we could also just manually find these values by performing functions on the dataset itself. The number of people is the number of rows:
```{r}
n <- nrow(data)
n
```

The minimum age is 
```{r}
min <- min(data$age)
min
```

whereas the maximum age is 
```{r}
max<- max(data$age)
max
```

We can use table function to quickly find the number of males and females
```{r}
table(data$sex)
```
Or we can use the dplyr function from the tidyverse package
```{r}
library(tidyverse)
data %>% count(sex)
```
We can perform similar functions for both the smoker and region subpopulations
```{r}
data %>% count(smoker)
data %>% count(region)
```
If we want to narrow it down to the number of men or number of women per region we can simply add on new arguments
```{r}
data %>% count(region,sex)
```

```{r}
# Count of individuals by sex and age group
data$age_group <- cut(data$age, breaks = c(17, 29, 40, 60, Inf), labels = c("18-29", "30-40", "41-60", "61+"))
demographic_counts <- table(data$sex, data$age_group)

# Visualize with a bar plot
barplot(demographic_counts, beside = TRUE, col = c("lightblue", "lightpink"),
        main = "Demographic Distribution by Age Group", xlab = "Age Group", ylab = "Count", legend = TRUE)



```


Similarily for men who are smokers with children
```{r}
data_men <- data %>% 
  filter(sex == "male")
data_men %>% count(children,smoker,)

```
We want to visualize a range of values for BMI thus let's using a histogram
```{r}
hist(data$bmi, col= "lightgreen", main="BMI Distribution", xlab = "BMI")
```

From this plot we can see that the highest frequencies are at a BMI centered around 30

Similarily for age:
```{r}
hist(data$age, col = "skyblue", main = "Age Distribution", xlab = "Age")

```


To find out how many male/females at each age:
```{r}
number_of_male_female <- data %>% count(age,sex)
number_of_male_female

```
Or we could do it based on age ranges
```{r}
# data %>% count(age %in% 18:25, sex)
age_groups <- data %>%
  mutate(age_group = case_when(
    age %in% 18:29 ~ '18-29',
    age %in% 30:40 ~ '30-40',
    age %in% 41:60 ~ '41-60',
    age >= 61 ~ '61+'
  )) %>%
  count(age_group, sex)

print(age_groups)

data_age_groups <- as.data.frame(age_groups)
```
Plotting the data onto a barplot:
```{r}
barplot(demographic_counts, beside = TRUE, col = c("lightblue", "lightpink"),
        main = "Number of Male and Females in each Age Group", xlab = "Age Group", ylab = "Count", legend = TRUE)


```

We can also use ggplot2 to plot the data out:
```{r}
library(ggplot2)
ggplot(data_age_groups, aes(x = age_group, y = n, fill = sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Males and Females in Each Age Group",
       x = "Age Group", y = "Count") +
  scale_fill_manual(values = c("lightblue", "lightpink"))
 
```

We want to see if there exists correlation between our variables so let's create a correlation matrix:
```{r}
library(corrplot)
cor_matrix <- cor(data[, c("age", "bmi", "children", "charges")])
corrplot(cor_matrix, method = "color")


```

```{r}
round(cor_matrix, 2)
```

From this matrix we can infer a few things:

1) All correlations are positive, but most are weak.

2) Age and charges have the strongest correlation, at about 0.30, followed by BMI and charges at about 0.20.

3) Age and BMI are barely correlated (about 0.11), and the number of children is almost unrelated to everything else.

These are simple pairwise correlations. They don't account for smoking, which turns out to matter far more than any of these variables (see below).


## Relationship between Smoking and Charges

To see the relationship between smoking and charges, we can compare summary statistics and perform a t-test:
```{r}
summary(data$charges[data$smoker == "yes"])
summary(data$charges[data$smoker == "no"])

```
From a quick glance we can see that on average, non-smokers paid less on average for insurance charges


```{r}
boxplot(data$charges ~ data$smoker, col = c("cyan", "red"),
        main = "Insurance Charges by Smoking Status", xlab = "Smoker", ylab = "Insurance Charges")
results<-t.test(data$charges[data$smoker == "yes"], data$charges[data$smoker == "no"])
print(results)
```

At first glance, our t-value is large, which suggests a large difference between the mean charges of smokers and non-smokers. Note that it is positive which implies that the smokers on average have a greater charge than non-smokers.
The p-value is far below 0.05, so smokers have significantly higher average charges than non-smokers. (This is Welch's t-test, which doesn't assume the two groups have equal variances. That matters here, since smokers' charges are much more spread out.)




```{r}
plot(data$bmi, data$charges, col = ifelse(data$smoker == "yes", "red", "blue"),
     pch = 16, main = "BMI vs. Insurance Charges by Smoking Status", xlab = "BMI", ylab = "Insurance Charges")



```

We can see from this scatterplot that the spread of BMI was roughly equal for smokers (red) and non-smokers (blue). The insurance charges for smokers was overall higher across the board, with a spike in distribution once BMI was greater than 30




```{r}
library(ggplot2)
ggplot(data, aes(x = smoker, y = charges)) +
  geom_violin(fill = "lightblue") +
  labs(title = "Insurance Charges by Smoking Status", x = "Smoker", y = "Insurance Charges")

```



## Relationship between BMI and charges

```{r}
summary(data$charges)
summary(data$bmi)
plot(data$bmi, data$charges, col = "blue", pch = 16,
     main = "BMI vs. Insurance Charges", xlab = "BMI", ylab = "Insurance Charges")
cor(data$bmi, data$charges)
library(ggplot2)
ggplot(data, aes(x = bmi, y = charges)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "BMI vs. Insurance Charges", x = "BMI", y = "Insurance Charges")
library(hexbin)
hb <- hexbin(data$bmi, data$charges, xbins = 20)
plot(hb, main = "Hexbin Plot: BMI vs. Insurance Charges", xlab = "BMI", ylab = "Insurance Charges")
cor.test(data$bmi, data$charges, method = "pearson")
lm_model <- lm(charges ~ bmi, data = data)
summary(lm_model)

```
Since our dataset is large, we can use a hexbin plot to further analyze the distribution. We can see that at a BMI of ~28-30, we have the highest density of people with insurance charges less than 10000. 

## Relationship between Age and Charges

Let's determine if there exists a relation between the age of an individual and the amount of their charges
```{r}
avg_charges_by_age <- tapply(data$charges, data$age, mean)

plot(names(avg_charges_by_age), avg_charges_by_age, type = "o", col = "blue",
     main = "Average Insurance Charges by Individual Age", xlab = "Age", ylab = "Average Charges")

```



We use an O plot here instead of a B plot to better see the trends

Note if we fit a trend line to the plot:
```{r}
plot(names(avg_charges_by_age), avg_charges_by_age, type = "o", col = "blue",
     main = "Average Insurance Charges by Individual Age", xlab = "Age", ylab = "Average Charges")
loess_model <- loess(avg_charges_by_age ~ as.numeric(names(avg_charges_by_age)))
lines(as.numeric(names(avg_charges_by_age)), predict(loess_model), col = "red")
```
Here, it is evident that there is a positive trend with charges regarding age. 

Let's generalize the data more into age groups instead of individual ages:
```{r}

avg_charges_by_age_group <- tapply(data$charges, data$age_group, mean)

barplot(avg_charges_by_age_group, col = "skyblue",
        main = "Average Insurance Charges by Age Group", xlab = "Age Group", ylab = "Average Charges")
```
Here, it is evident the older age groups spend more on charges than the groups previous to them
















































































Suppose we want to predict what the insurance charges would be based on one's BMI alone.
```{r}
pred_charge <- lm(charges ~ bmi, data = data)
summary(pred_charge)
```

BMI alone explains only about 4% of the variation in charges (R-squared). Adding sex makes almost no difference:
```{r}
summary(lm(charges ~ bmi + sex, data = data))$adj.r.squared
```

## Relationship between Region and Charges

Let us see if there exists a relation between the region an individual inhabits and the charges they possess
```{r}
region_counts <- table(data$region)

pie(region_counts, labels = names(region_counts), main = "Distribution by Region")

```

From this pie chart, we can see each region roughly holds an equivalent portion

```{r}
avg_charges_by_region <- tapply(data$charges, data$region, mean)

barplot(avg_charges_by_region, col = c("lightblue", "lightgreen", "lightpink", "lightyellow"),
        main = "Average Insurance Charges by Region", ylab = "Average Charges")

```
At a quick glance, the southeast region seems to have the highest average charges, followed by the northeast and finally the northwest and southwest. To get further information, let's perform an ANOVA test
```{r}


anova_result <- aov(charges ~ region, data = data)

summary(anova_result)


```
Note here out p value is less than 0.05 and our F-value is 2.97, hence this suggests that we reject our null hypothesis of the means among the regions being equal and that there exists a significant difference in means between at least one of the regions

```{r}

tukey_results <- TukeyHSD(anova_result)
print(tukey_results)

```
We can see from our Tukey test that the adjusted p-value for the southwest vs southeast comparison is just below 0.05 (0.048), so those means are significantly different. Southeast vs northwest is just above (0.058), and the other pairs are far from significant.

Charges are very skewed and far more spread out for smokers, which strains ANOVA's equal-variance assumption, so treat these p-values as approximate. More importantly, the regions differ in who lives there:

```{r}
data %>%
  group_by(region) %>%
  summarise(share_smokers = mean(smoker == "yes"), mean_bmi = mean(bmi), mean_charges = mean(charges))
```

The southeast has the highest share of smokers and the highest average BMI. That alone could explain its higher average charges, which we check in the regression below.


```{r}
demographic_counts_by_region <- table(data$region, data$sex)

col_palette <- c("skyblue", "lightcoral", "lightgreen", "lightsalmon")

barplot(demographic_counts_by_region, beside = TRUE, col = col_palette,
        main = "Demographic Distribution by Region", xlab = "Region", ylab = "Count", legend = TRUE)

```

```{r}
smoking_counts_by_region <- table(data$region, data$smoker)

col_palette <- c("lightblue", "lightgreen", "lightcoral", "lightsalmon")

barplot(smoking_counts_by_region, beside = TRUE, col = col_palette,
        main = "Smoking Distribution by Region", xlab = "Region", ylab = "Count", legend = TRUE)


```

```{r}
library(corrplot)
data_by_region <- split(data, data$region)

correlation_matrices <- lapply(data_by_region, function(subdata) cor(subdata[, c("age", "bmi", "children", "charges")]))

lapply(correlation_matrices, function(cor_matrix) corrplot(cor_matrix, method = "color"))

```

## Predicting Charges of an Individual

Given the age,sex,BMI, number of children, if they are a smoker, and the region they are from, we want to be able to predict the charges that an individual would have. We will perform regression analysis:

```{r}
# Here we are converting variables sex and smoker into factors
data$sex <- as.factor(data$sex)
data$smoker <- as.factor(data$smoker)

data2 <- cbind(data, model.matrix(~ region - 1, data = data))
model <- lm(charges ~ age + sex + bmi + children + smoker + region, data = data2)
summary(model)
```
Our intercept estimate is -11938.5, we can ignore this since this is the expected value when all other variables are 0 hence this is redundant.
The age estimate is 256.9 which means that the charges are estimated to be increased by 256.9 for every additional unit of age (year).
The sexmale estimate is -131.3 which would normally suggest that the charges for males are lower but since our p-value is greater than 0.05, we can conclude this variable is not statistically significant.
The BMI estimate is 339.2 which means that for every unit increase in BMI, the charges increase by 339.2.
The children estimate is 475.5 which means for every additional child an individual has, the charges increases by 475.5.
For the smokeryes variable, this has an estimate of 23848.5 with a very small p-value. This means that if an individual is a smoker, then their charges are estimated to increase by 23848.5, by far the largest margin of the variables so far.

We can see that the southeast region has an estimate of -1035 compared to our northeast region, which seems to contradict our previous barplot. It doesn't: the barplot compared raw averages, while the regression compares people with the same age, BMI, children and smoking status. The southeast's higher raw average comes from its higher share of smokers and higher BMI (confounding). Once those are held constant, living in the southeast is associated with slightly lower charges.

Next, we check the model's assumptions: normality of residuals and constant variance (homoscedasticity).

```{r}

qqnorm(residuals(model))
qqline(residuals(model))


plot(fitted(model), residuals(model), col = "blue", pch = 16,
     main = "Residuals vs. Fitted", xlab = "Fitted Values", ylab = "Residuals")

```
We can also try removing variables that are not significant in our regression. Note that region is one variable with three dummy columns, so it should be tested as a whole rather than column by column. An F-test comparing the models with and without sex and region does this:

```{r}
model_new <- lm(charges ~ age + bmi + children + smoker, data = data2)
summary(model_new)
anova(model_new, model)
```

The F-test's p-value is above 0.05, so dropping sex and region together doesn't make this model significantly worse.
```{r}
qqnorm(residuals(model_new))
qqline(residuals(model_new))


plot(fitted(model_new), residuals(model_new), col = "blue", pch = 16,
     main = "Residuals vs. Fitted", xlab = "Fitted Values", ylab = "Residuals")

```
From our Q-Q plot, we can see that the residuals are not normal. The residuals vs fitted plot shows separate bands and a large gap rather than a random cloud. That pattern means the model is missing something, not just that the errors are unusual. One common response is to transform charges:

```{r}
data2$log_charges <- log(data2$charges + 1)  # we add 1 here to handle zero values
model_transformed <- lm(log_charges ~ age + bmi + children + smoker, data = data2)
summary(model_transformed)


```
Our new Q-Q plot:

```{r}
residuals_log_model <- residuals(model_transformed)
qqnorm(residuals_log_model)
qqline(residuals_log_model)

```
We can still see deviation from normality even with the log transform.

```{r}
library(e1071)  
skewness(residuals_log_model)
kurtosis(residuals_log_model)


```
A skewness of about 1.6 and an excess kurtosis of about 4.2 mean the residuals are not symmetric and have heavy tails. Robust regression (`rlm`) is one response. It is not a transformation: it gives less weight to observations with large residuals, so a few extreme charges have less influence on the fit:

```{r}
library(MASS)

model_robust <- rlm(log_charges ~ age + bmi + children + smoker, data = data2)
summary(model_robust)
plot(model_robust)



```

### Fixing the model rather than the residuals

Transforming charges and downweighting outliers treat the symptoms. The bands in the residual plot come from a missing interaction: smoking raises charges far more for people who are obese (BMI 30 or more) than for those who aren't. The earlier scatterplot showed this, with smokers' charges jumping once BMI passed 30. Charges also rise faster than linearly with age.

```{r}
ggplot(data, aes(x = bmi, y = charges, colour = smoker)) +
  geom_point(alpha = 0.6) +
  geom_vline(xintercept = 30, linetype = "dashed") +
  labs(title = "Smoking raises charges much more above BMI 30", x = "BMI", y = "Insurance Charges")
```

Adding an obesity indicator, its interaction with smoking, and a squared age term:

```{r}
data2$obese <- data2$bmi >= 30
model_interaction <- lm(charges ~ age + I(age^2) + children + bmi + smoker * obese + sex + region, data = data2)
summary(model_interaction)

c(original = summary(model)$adj.r.squared, with_interaction = summary(model_interaction)$adj.r.squared)
```

Adjusted R-squared rises from about 0.75 to about 0.87. The `smokeryes:obeseTRUE` estimate is the extra charge for smokers who are also obese, roughly 20,000 on top of the smoking effect.

```{r}
plot(fitted(model_interaction), residuals(model_interaction), col = "blue", pch = 16,
     main = "Residuals vs. Fitted (interaction model)", xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, lty = 2)
```

Most residuals now sit in one band around zero. The points well above it are a minority of non-smokers whose charges are far higher than anything in the data explains, such as an illness the dataset doesn't record. No transformation can fix that. It's a limit of the available variables.

## Prediction of an Individual's charges example

Here we will do a quick example with the interaction model, for a 34-year-old male smoker from the northeast with a BMI of 27 and no children:
```{r}
new_data <- data.frame(age = 34, sex = "male", bmi = 27, children = 0, smoker = "yes",
                       region = "northeast", obese = FALSE)
predict(model_interaction, newdata = new_data, interval = "prediction", level = 0.95)
```

The model predicts charges of about 19,500, with a 95% prediction interval of roughly 10,700 to 28,300. The interval is wide because individual charges vary a lot even among similar people. As a check, smokers aged 30-38 with a BMI under 30 in the data have median charges of about 19,400:
```{r}
data %>% filter(smoker == "yes", bmi < 30, age >= 30, age <= 38) %>% summarise(n = n(), median_charges = median(charges))
```

(An earlier version predicted from the robust model on the log scale and reported `exp()` of the prediction as the expected charge. That has two problems. The model was fitted to log(charges + 1), so the back-transform is `exp(prediction) - 1`. And exponentiating a predicted log value gives roughly the median charge, not the mean. `predict()` also doesn't produce valid prediction intervals for `rlm` models.)




```{r}
# Check for missing/infinite values in 'age' and 'charges'
any(!is.finite(data2$age))
any(!is.finite(data2$charges))
```


## Interactive histogram

This app needs a live R session: run this chunk in RStudio. It is not evaluated in the static HTML.

```{r, eval = FALSE}
library(shiny)
library(dplyr)




ui <- fluidPage(
  titlePanel("Interactive Histogram"),
  sidebarLayout(
    sidebarPanel(
      selectInput("variable", "Choose Variable:",
                  choices = c("age", "bmi"),
                  selected = "age"),
      # The range slider is rebuilt for whichever variable is selected
      uiOutput("range_slider")
    ),
    mainPanel(
      plotOutput("histogram")
    )
  )
)


server <- function(input, output) {
  # The slider's limits follow the selected variable (age 18-64, BMI about 16-53).
  # Before, the slider always used age limits, so choosing BMI filtered on age-sized numbers.
  output$range_slider <- renderUI({
    values <- data[[input$variable]]
    sliderInput("range", "Select Range:",
                min = floor(min(values)), max = ceiling(max(values)),
                value = c(floor(min(values)), ceiling(max(values))), step = 1)
  })

  # Reactive expression to filter data based on user inputs
  filtered_data <- reactive({
    req(input$range)
    data %>%
      filter(get(input$variable) >= input$range[1],
             get(input$variable) <= input$range[2])
  })


  output$histogram <- renderPlot({
    if (length(unique(filtered_data()$charges)) > 1) {
      hist(filtered_data()$charges, main = "Distribution of Charges",
           xlab = "Charges", col = "lightblue", border = "black")
    } else {
      plot(1, type = "n", axes = FALSE, xlab = "", ylab = "")
      text(1, 1, "Not enough data for a histogram.", cex = 1.5, col = "red")
    }
  })
}


shinyApp(ui = ui, server = server)


```








```


























































































