---
title: "STAC51 A2"
output: pdf_document
date: "2024-03-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

install.packages("binom")



### Q2a)
```{r}
set.seed(2024)
n11 <- rbinom(1, 10, 0.4)
n21 <- rbinom(1, 10, 0.4)
print(n11)
print(n21)

```

### Q2b)
```{r}
n1plus <- 10
n2plus <- 10
ntable <- array(data = c(n11, n21, n1plus - n11, n2plus - n21), dim = c(2,2))
print(ntable)
```

### Q2c)
```{r}
fisher_result <- fisher.test(ntable)
print(fisher_result)
```
Our p-value is 0.3698. Since this is greater than 0.05 for 5 percent level of signif, we fail to reject the null hypothesis.


### Q2d)
```{r}
set.seed(2024)
p_val <- replicate(10000, {
n11 <- rbinom(1, 10, 0.4) 
n21 <- rbinom(1, 10, 0.4)
ntable <- array(data = c(n11, n21, n1plus - n11, n2plus - n21), dim = c(2,2))
fisher.test(ntable)$p.val})
print(head(p_val, 5)) 
rej <- sum(p_val < 0.05)
print(rej)

```
We rejected the null hypothesis 142 times out of 10000, which is roughly 1.42%. This was for the 5 percent level of significance. Our rate of 1.42% is lower than the expected rate of 5%.  


### Q2e)
```{r}
set.seed(2024)
n1plus <- 100
n2plus <- 100
p_vals <- replicate(10000, {
n11 <- rbinom(1, 100, 0.4)
n21 <- rbinom(1, 100, 0.4)
ntable <- array(data = c(n11, n21, n1plus - n11, n2plus - n21), dim = c(2,2))
fisher.test(ntable)$p.value
})
print(head(p_vals, 5)) # Print the first five p-values
rej <- sum(p_vals < 0.05)
print(rej)

```
We rejected the null hypothesis 358 times out of 10000. This is roughly 3.58%. This was for the level of signif of 5%. This was closer to our expected rejection rate of 5% as it was higher than the one from d) but still lower than expected.


### Q3j)
```{r}
cases_exposed <- 3
cases_not_exposed <- 1
controls_exposed <- 1
controls_not_exposed <- 3

n_cases <- cases_exposed + cases_not_exposed
n_controls <- controls_exposed + controls_not_exposed
eta_null <- 0.1

# Exact conditional test of H0: log odds ratio = eta_null.
# Conditional on the total number exposed (t), the number of exposed cases X follows a
# noncentral hypergeometric distribution: P(X = x) is proportional to
# choose(n_cases, x) * choose(n_controls, t - x) * exp(eta_null * x).
# (The earlier version gave both groups the same success probability, so eta_null cancelled out,
#  and it summed joint probabilities without dividing by P(X + Y = t), so the result wasn't a p-value.)
t <- cases_exposed + controls_exposed
x_values <- max(0, t - n_controls):min(n_cases, t)
weights <- choose(n_cases, x_values) * choose(n_controls, t - x_values) * exp(eta_null * x_values)
probs <- weights / sum(weights)
p_observed <- probs[x_values == cases_exposed]

p_value_two_sided <- sum(probs[probs <= p_observed + 1e-12])   # outcomes no more likely than the observed one
p_value_upper <- sum(probs[x_values >= cases_exposed])         # one-sided: as many or more exposed cases
c(two_sided = p_value_two_sided, one_sided_upper = p_value_upper)
```
The two-sided p-value is about 0.49 (one-sided about 0.27). Both are well above 0.05, so we fail to reject the null hypothesis that the log odds ratio is 0.1. With only 8 subjects, the data can't distinguish this odds ratio from many others.

### Q3k)
```{r}
cases_exposed <- 3
cases_not_exposed <- 1
controls_exposed <- 1
controls_not_exposed <- 3

n_cases <- cases_exposed + cases_not_exposed
n_controls <- controls_exposed + controls_not_exposed

# Conditional likelihood of the log odds ratio eta, given the total number exposed:
# L(eta) = choose(n1, x) choose(n0, t - x) exp(eta x) / sum over u of choose(n1, u) choose(n0, t - u) exp(eta u)
# (The earlier version multiplied two binomials with the same probability, which peaks at eta = 0
#  for any 2x2 table, so it always estimated an odds ratio of 1.)
conditional_likelihood <- function(eta, n_cases, n_controls, cases_exposed, total_exposed = 4) {
  u <- max(0, total_exposed - n_controls):min(n_cases, total_exposed)
  weights <- choose(n_cases, u) * choose(n_controls, total_exposed - u) * exp(eta * u)
  weights[u == cases_exposed] / sum(weights)
}
etas <- seq(-3, 3, 0.01)
likelihoods <- sapply(etas, conditional_likelihood, n_cases, n_controls, cases_exposed)
max_likelihood_eta <- etas[which.max(likelihoods)]
max_likelihood_theta <- exp(max_likelihood_eta)

plot(etas, likelihoods, type = "l", xlab = "eta", ylab = "Likelihood",
     main = "Conditional Likelihood of eta")
points(max_likelihood_eta, max(likelihoods), col = "red", pch = 20)
max_likelihood_theta


```


















