---
title: "STAC51 A1"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
date: "2024-02-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
install.packages("binom")
install.packages("rootSolve")
## Question 1

#a)

Seasons is nominal (no inherent order or ranking)

#b)

Number of typos is ratio (ordered with consistent intervals and true zero point)

#c)

Employees rating of satisfaction is ordinal (meaningful order but intervals between them are not consistent)

#d) 

College student's SAT score is ratio (ordered with consistent intervals and true zero point)

#e)

Person's daily average level of activity is ordinal (meaningful order but intervals between them are not consistent)

##Question 2

```{r}
library(binom)
library(rootSolve)
n <- 35
y <- 28

# (a) 
wald_ci <- binom.confint(y, n, conf.level = 0.95, methods = "asymptotic")
wald_ci

alpha <- 0.05

# (b) 
ac_ci <- binom.confint(y, n, conf.level = 0.95, methods = "ac")
ac_ci



# (c) 
likelihood_ratio_ci <- binom.confint(y, n, conf.level = 0.95, methods = "lrt")
likelihood_ratio_ci

# (d)
clopper_pearson_ci <- binom.confint(y, n, methods = "exact", 1-alpha)
clopper_pearson_ci



```

## Question 3

```{r}
# (a) 
n <- 30
p_employed_study <- 0.65
p_employed_unrelated <- 0.25
p_unemployed <- 0.10

probability_a <- choose(n, 17) * choose(n - 17, 10) * (p_employed_study^17) * (p_employed_unrelated^10) * (p_unemployed^(n - 17 - 10))
probability_a

# (b) 
k_unemployed <- 3
probability2 <- choose(n, k_unemployed) * (p_unemployed^k_unemployed) * ((1 - p_unemployed)^(n - k_unemployed))
probability2

# (c) Calculate the conditional probability
probability3 <- sum(dbinom(1:n, size = n, prob = p_employed_study) * dbinom(0:(n-1), size = n-1, prob = p_employed_unrelated)) / sum(dbinom(1:n, size = n, prob = p_employed_study) * dbinom(0:(n-1), size = n-1, prob = p_employed_unrelated))
probability3


```
## Question 4

```{r}
# (a)i 
n <- 40
pi <- 0.05
x_values <- 0:4

probability_exact <- pbinom(max(x_values), size = n, prob = pi)
probability_exact

# (a)ii 
lambda <- n * pi
probability_poisson <- ppois(max(x_values), lambda)
mean_normal <- n * pi
sd_normal <- sqrt(n * pi * (1 - pi))
probability_normal <- pnorm(max(x_values) + 0.5, mean_normal, sd_normal)  # Continuity correction

result_comparison <- data.frame(
  Method = c("Exact (Binomial)", "Poisson Approximation", "Normal Approximation"),
  Probability = c(probability_exact, probability_poisson, probability_normal)
)

result_comparison

# (b)i 
pi_b <- 0.45
x_values_b <- 0:10

probability_exact_b <- pbinom(max(x_values_b), size = n, prob = pi_b)
probability_exact_b

# (b)ii 
lambda_b <- n * pi_b
probability_poisson_b <- ppois(max(x_values_b), lambda_b)
mean_normal_b <- n * pi_b
sd_normal_b <- sqrt(n * pi_b * (1 - pi_b))
probability_normal_b <- pnorm(max(x_values_b) + 0.5, mean_normal_b, sd_normal_b)  # Continuity correction

result_comparison_b <- data.frame(
  Method = c("Exact (Binomial)", "Poisson Approximation", "Normal Approximation"),
  Probability = c(probability_exact_b, probability_poisson_b, probability_normal_b)
)

result_comparison_b

```

## Question 5

```{r}

y <- c(53, 67, 58, 65, 69, 76, 59, 55, 66, 67)


n <- length(y)
mu <- 65


likelihood_function <- function(v) {
  a <- n / 2
  b <- sum((y - mu)^2) / 2
  return(v^(-a) * exp(-b / v))
}


v_values <- seq(25, 75, length.out = 1000)
likelihood_values <- sapply(v_values, likelihood_function)
max_likelihood_estimate <- v_values[which.max(likelihood_values)]

plot(v_values, likelihood_values, type = "l", col = "blue", lwd = 2,
     xlab = "v", ylab = "L(v)", main = "Likelihood Function")

abline(v = max_likelihood_estimate, col = "red", lty = 2)
points(max_likelihood_estimate, likelihood_function(max_likelihood_estimate), col = "red", pch = 16)

cat("Maximum Likelihood Estimate of v:", max_likelihood_estimate, "\n")



```



## Question 6

```{r}
#a)

n <- 14
pi <- 0.4
alpha <- 0.05


critical_region <- qbinom(alpha, size = n, prob = pi, lower.tail = TRUE)
critical_region

#b)
type_I_error_probability <- pbinom(critical_region, size = n, prob = pi, lower.tail = TRUE)
type_I_error_probability


```


## Question 7

```{r}
# Part (a)
n_a <- 10
p_a <- 0.25
x_a <- 0:n_a

hist_data_a <- dbinom(x_a, size = n_a, prob = p_a)
barplot(hist_data_a, names.arg = x_a, col = "skyblue", xlab = "Number of Successes", ylab = "Probability",
        main = "Binomial Distribution (n = 10, p = 0.25)")

# Part (b)
n_b <- 10
p_b <- 0.5
x_b <- 0:n_b

hist_data_b <- dbinom(x_b, size = n_b, prob = p_b)
barplot(hist_data_b, names.arg = x_b, col = "lightgreen", xlab = "Number of Successes", ylab = "Probability",
        main = "Binomial Distribution (n = 10, p = 0.5)")

# Part (c)
n_c <- 10
p_c <- 0.75
x_c <- 0:n_c

hist_data_c <- dbinom(x_c, size = n_c, prob = p_c)
barplot(hist_data_c, names.arg = x_c, col = "lightcoral", xlab = "Number of Successes", ylab = "Probability",
        main = "Binomial Distribution (n = 10, p = 0.75)")

# Part (d)
n_d <- 100
p_d <- 0.75
x_d <- 0:n_d

hist_data_d <- dbinom(x_d, size = n_d, prob = p_d)
barplot(hist_data_d, names.arg = x_d, col = "lightpink", xlab = "Number of Successes", ylab = "Probability",
        main = "Binomial Distribution (n = 100, p = 0.75)")



```

## Question 8

```{r}
# Part (a)
lambda_a <- 2.5
x_a <- 0:7

hist_data_a <- dpois(x_a, lambda_a)
barplot(hist_data_a, names.arg = x_a, col = "lightblue", xlab = "Number of Events", ylab = "Probability",
        main = "Poisson Distribution (λ = 2.5)")

# Part (b)
lambda_b <- 20
x_b <- 7:33

hist_data_b <- dpois(x_b, lambda_b)
barplot(hist_data_b, names.arg = x_b, col = "lightgreen", xlab = "Number of Events", ylab = "Probability",
        main = "Poisson Distribution (λ = 20)")


```


