---
title: "STAC51 A2"
output: pdf_document
date: "2024-03-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

install.packages("binom")



### Q2a)
```{r}
set.seed(2024)
n11 <- rbinom(1, 10, 0.4)
n21 <- rbinom(1, 10, 0.4)
print(n11)
print(n21)

```

### Q2b)
```{r}
n1plus <- 10
n2plus <- 10
ntable <- array(data = c(n11, n21, n1plus - n11, n2plus - n21), dim = c(2,2))
print(ntable)
```

### Q2c)
```{r}
fisher_result <- fisher.test(ntable)
print(fisher_result)
```
Our p-value is 0.3698. Since this is greater than 0.05 for 5 percent level of signif, we fail to reject the null hypothesis.


### Q2d)
```{r}
set.seed(2024)
p_val <- replicate(10000, {
n11 <- rbinom(1, 10, 0.4) 
n21 <- rbinom(1, 10, 0.4)
ntable <- array(data = c(n11, n21, n1plus - n11, n2plus - n21), dim = c(2,2))
fisher.test(ntable)$p.val})
print(head(p_val, 5)) 
rej <- sum(p_val < 0.05)
print(rej)

```
We rejected the null hypothesis 142 times out of 10000, which is roughly 1.42%. This was for the 5 percent level of significance. Our rate of 1.42% is lower than the expected rate of 5%.  


### Q2e)
```{r}
set.seed(2024)
n1plus <- 100
n2plus <- 100
p_vals <- replicate(10000, {
n11 <- rbinom(1, 100, 0.4)
n21 <- rbinom(1, 100, 0.4)
ntable <- array(data = c(n11, n21, n1plus - n11, n2plus - n21), dim = c(2,2))
fisher.test(ntable)$p.value
})
print(head(p_vals, 5)) # Print the first five p-values
rej <- sum(p_vals < 0.05)
print(rej)

```
We rejected the null hypothesis 358 times out of 10000. This is roughly 3.58%. This was for the level of signif of 5%. This was closer to our expected rejection rate of 5% as it was higher than the one from d) but still lower than expected.


### Q3j)
```{r}
cases_exposed <- 3
cases_not_exposed <- 1
controls_exposed <- 1
controls_not_exposed <- 3

n_cases <- cases_exposed + cases_not_exposed
n_controls <- controls_exposed + controls_not_exposed
eta_null <- 0.1

binom_prob <- function(x, n, p) {
  choose(n, x) * p^x * (1 - p)^(n - x)
}

likelihood_null <- binom_prob(cases_exposed, n_cases, exp(eta_null) / (1 + exp(eta_null))) * 
                   binom_prob(controls_exposed, n_controls, exp(eta_null) / (1 + exp(eta_null)))
p_value <- 0
for (x in 0:n_cases) {
  for (y in 0:n_controls) {
    if (x + y == cases_exposed + controls_exposed) {
      likelihood <- binom_prob(x, n_cases, exp(eta_null) / (1 + exp(eta_null))) * 
                    binom_prob(y, n_controls, exp(eta_null) / (1 + exp(eta_null)))
      
      
      if (likelihood >= likelihood_null) {
        p_value <- p_value + likelihood
      }
    }
  }
}
p_value
```
Our p-value here is 0.26298 which is greater than our null hypothesis value of 0.1, hence we reject the null hypothesis here.

### Q3k)
```{r}
cases_exposed <- 3
cases_not_exposed <- 1
controls_exposed <- 1
controls_not_exposed <- 3

n_cases <- cases_exposed + cases_not_exposed
n_controls <- controls_exposed + controls_not_exposed

conditional_likelihood <- function(eta, n_cases, n_controls, cases_exposed) {
  p <- exp(eta) / (1 + exp(eta))
  likelihood_cases <- dbinom(cases_exposed, n_cases, p)
  likelihood_controls <- dbinom(n_controls - cases_exposed, n_controls, p)
  return(likelihood_cases * likelihood_controls)
}
etas <- seq(-3, 3, 0.01)
likelihoods <- sapply(etas, conditional_likelihood, n_cases, n_controls, cases_exposed)
max_likelihood_eta <- etas[which.max(likelihoods)]
max_likelihood_theta <- exp(max_likelihood_eta)

plot(etas, likelihoods, type = "l", xlab = "eta", ylab = "Likelihood",
     main = "Conditional Likelihood of eta")
points(max_likelihood_eta, max(likelihoods), col = "red", pch = 20)
max_likelihood_theta


```


















